---
title: "Workshop2"
author: "Everett"
format: html
editor: visual
---

# 2.12

This project is about optimizing and consolidating multiple datasets into a clean, finished product using commands in RStudio (obviously, all of these workshops are in RStudio).

## Phase 1

Phase 1 is our setup. In order to analyze the data, we need to clean up the mess we've been handed. Before anything else, we need to populate the library with the necessary packages. Consider this our preamble/staging step.

```{r packages echo = FALSE}
library(tidyverse) # Ole reliable, this includes a load of useful sub-programs: dplyr, ggplot2, lubridate (dates, as the name would suggest), and stringr (text).
library(readxl) # One of our datesets is stored in an excel rather than a csv file, so we'll need a bespoke package to read and manipulate it.
```

### 1. Import the .csv files

Now that our packages are installed, let's load all of the .csv files into our library. We won't do anything with them for now, but I find it's best to load them all in a single block that can be run whenever we re-open the project.

```{r load-data}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
source(here::here("code/diw_sensor.R"))
```

### 2. Prepping estuary_catch_log.xlsx

The excel file has its data spread across four tabs, so we'll need to load each tab individually so that we can view and manipulate them in RStudio.

```{r load-catch-1}
catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = 1) # You can alternatively input the name of each tab rather than its number, but opting for the number is faster if you don't know these names.
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)
```

Check your environment to ensure that everything loaded properly. Now it's time to combine these sheets into a single file.

```{r bind-catch}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
rm(catch1, catch2, catch3, catch4)
```

Now that we've run this, we have created a single data file named "catch_data" that holds everything from the excel document. We'll check to see how many levels we have for locations and species, and in so doing make sure that these levels are consistently named within our combined dataset.

```{r distinct-catch}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```

Because we're compartmentalizing our data wrangling, we'll pick a notation and just apply it using stringr and lubridate as we get to each dataset. For this project, we'll make everything lowercase and use underscores instead of spaces.

```{r fix-site-names}
catch_data <- catch_data |>
  mutate(
  # Get rid of leading and trailing spaces
    site = str_trim(site),
  # Make everything lowercase
    site = str_to_lower(site),
  # Turn the spaces we want into underscores
    site = str_replace_all(site, " ", "_"),
  # Do the same for the species entries
    species = str_trim(species),
  # Make everything lowercase
    species = str_to_lower(species),
  # Turn the spaces we want into underscores
    species = str_replace_all(species, " ", "_"))|> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
  # Make sure everything looks correct
print(catch_data)

```

### 3. Standardize our other data

Now that we've set our syntax standards on our catch_data file, it will serve as the nexus to which we'll add data from our other files using the join() function. To save us a headache, we're going to apply the same syntax (all lower case and underscore separated) to each file before we begin the great joining. Let's start with ensuring that the site names in the metadata match our catch_data.
```{r inspect-metadata}
head(metadata)
glimpse(metadata)
```
Decidedly not. Let's fix that by running the same code framework we applied to catch_data.
```{r fix-site-names}
metadata <- metadata |>
  mutate(
    # By using additional parenthetical, you can actually nest the trim command within the command to make everything lower case
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all characters to factors, this will help us greatly when it's time to wrangle our data
  mutate(across(where(is.character), as_factor)
  )
```
Now run our glimpse code again to make sure it worked...
```{r inspect-metadata}
head(metadata)
glimpse(metadata)
```
... and it did!